
---
title: "Exploratory Data Analysis on Loans"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

by KRISTIJAN BAKARIC
========================================================

Introduction to the dataset that EDA will be carried on: 
 
Last updated 03/11/2014

This data set contains 113,937 loans with 81 variables on each loan, including 
loan amount, borrower rate (or interest rate), current loan status, 
borrower income, borrower employment status, borrower credit history,
and the latest payment information.

There are MANY variables in this data set and you are not 
expected to explore all of them. You should explore between 
10-15 variables in your analysis.
Expected time to analyze the dataset 15-30 hours

**Since the dataset is to vast to explore in this excercise I will 
intimately examine only 10-15 variables and create a dataframe from 
the relevant selection.**

This variable dictionary explains the variables in the subset of the data set
that I will explore:

* listing keys; Unique key for each listing, 
same value as the 'key' used in the listing object in the API.

* BorrowerState: The two letter abbreviation of the state of the address of the 


* Term: The length of the loan expressed in months.


* LoanStatus: The current status of the loan: Cancelled,  Chargedoff, Completed, 
Current, Defaulted, FinalPaymentInProgress, PastDue. 
The PastDue status will be accompanied by a delinquency bucket.
 
* LoanOriginalAmount: The origination amount of the loan.

* ClosedDate: Closed date is applicable for Cancelled, Completed, Chargedoff 
and Defaulted loan statuses. 


* BorrowerRate: The Borrower's interest rate for this loan.


* ListingCategory..numeric.: The category of the listing that the borrower 
selected when posting their 
listing: 0 - Not Available, 1 - Debt Consolidation, 2 - Home Improvement, 
3 - Business, 4 - Personal Loan, 5 - Student Use, 6 - Auto, 7- Other, 
8 - Baby&Adoption, 9 - Boat, 10 - Cosmetic Procedure, 
11 - Engagement Ring, 12 - Green Loans, 13 - Household Expenses, 
14 - Large Purchases, 15 - Medical/Dental, 16 - Motorcycle, 
17 - RV, 18 - Taxes, 19 - Vacation, 20 - Wedding Loans


* Occupation: The Occupation selected by the Borrower at the time they created 
  the listing.


* EmploymentStatus: The employment status of the borrower at the time they 
  posted the listing.


* IsBorrowerHomeowner: A Borrower will be classified as a homowner if they 
have a mortgage on their credit profile or provide 
documentation confirming they are a homeowner.


* IncomeRange: The income range of the borrower at the time the listing 
 was created.


* DebtToIncomeRatio_ The debt to income ratio of the borrower at the time 
the credit profile was pulled. This value is Null if the 
debt to income ratio is not available. This value is 
capped at 10.01 (any debt to income ratio larger than
1000% will be returned as 1001%).


* IncomeVerifiable: The borrower indicated they have the required documentation 
to support their income.

* LoanOriginationDate- date the loan was originated


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.
#install.packages("ggplot2", dependencies = T) 
#install.packages("knitr", dependencies = T)
#install.packages("dplyr", dependencies = T)
library(knitr)
library(ggplot2)
library(dplyr)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
## In the following section I will load the data, subset the data to a managable
## variable size, inspect briefly subset variables
## and will join the states name with state codes

## Load the Dataset
df <- read.csv("C:\\Users\\Bruker\\edukacija\\analytics_term_2\\eda\\project1\\prosperLoanData.csv",na.strings=c("NA","NaN", " ",""))

## load lookup table for country codes                        
stateslookup <- read.csv("C:\\Users\\Bruker\\edukacija\\analytics_term_2\\eda\\project1\\us_states2.csv")

## there is a mistmatch ofnumber of unique values in country codes in dataset
## and the number of country tables in the lookup table
#length(unique(df$BorrowerState))
#length(unique(stateslookup$StateCode))  

### attach full country names to the state codes
df <- df %>% 
  left_join(stateslookup, by = c("BorrowerState" = "StateCode"))


### subset the original dataframe to 10-15 variables

cols <- c("ListingKey", "LoanOriginalAmount","BorrowerState","Term","LoanStatus"
          , "ClosedDate", "BorrowerRate","ListingCategory..numeric.", 
          "Occupation", "EmploymentStatus", "IsBorrowerHomeowner","IncomeRange",
          "DebtToIncomeRatio", "StateCapital" , "LoanOriginationDate")

df_s <- df %>% 
  select(cols)

### inspect the head of the subset of the dataset for the EDA
head(df_s)

```

# Univariate Plots Section

```{r echo=FALSE, fig.width=10, fig.height=6,  message=FALSE, warning=FALSE}
### inspect the data health of the selected columns
#install.packages("ggthemes")
#install.packages("scales")
#install.packages("Gmisc")
#install.packages("funModeling")
library("ggthemes")
library("scales")
library(dplyr) ### library used for creating an additional derived metric 
              ###"DataCompletePerc"
library(funModeling) ### library utilized for deriving metrics dataframe


### before any data exploritory analysis I like to identify data completnees 
### report and visualize it utilizing funModeling library

DataFrameMetrics<- df_status (df_s)
DataFrameMetrics<- DataFrameMetrics %>% 
  mutate(DataCompletePerc=(p_na-100)*-1) ### % per column that has non-NA data

ggplot(DataFrameMetrics,aes(reorder(variable,p_na),
                            DataCompletePerc,fill=DataCompletePerc)) +
   geom_bar(stat="identity")+
   theme_hc() +
   scale_colour_hc()+
   theme(axis.text.x=element_text(angle=45,hjust=1))+
   guides(fill=FALSE) +
   theme(axis.text=element_text(size=14),
   axis.title=element_text(size=20,face="bold")) +
   ggtitle("") +
   xlab("Column Names") + ylab("Columns complete [%]")+
   geom_hline(yintercept = 50, color="red",size=1.5) +
   theme( # remove the vertical grid lines
   panel.grid.major.x = element_line( size=.1, color="black" ) ,
      # explicitly set the horizontal lines (or they will disappear too)
      panel.grid.major.y = element_line( size=.1, color="black" )
   )

```

From the variables I will explore, BorrowerState, Occupation, EmploymentStatus,
DebtToIncomeRatio and State Capital have percentage of missing observations 
in magnitude of up to 7.5%. CLosedDate for the loans has 
the biggest loss of observations, 51.6%


```{r echo=FALSE, Sumary}
summary(df_s)
```


```{r echo=FALSE, fig.width=10, fig.height=6, Univariate_Plots}

summary(df_s$LoanOriginalAmount)

#plot
ggplot(data=df_s, aes(df_s$LoanOriginalAmount)) + 
  geom_histogram(breaks=seq(0, max(df_s$LoanOriginalAmount), by = 1000), 
                 col="black", 
                 alpha = .6,
                 aes(fill=..count..)) + 
  labs(title="Histogram for Loan Amount") +
  labs(x="Loan Amount [USD]", y="Count") +
  scale_fill_gradient("Count", low = "green", high = "red") 

#plot
#grid.arrange( ggplot(aes(x=LoanOriginalAmount), 
 #       data = df_s) +
  #geom_histogram( bins = 30) ,
   #ggplot(aes(x=1, y=LoanOriginalAmount), 
    #    data = df_s) +
  #geom_boxplot( )  , nrow =1)

```

Distribution of Original loan amounts is right tailed and it has 5 distinct
peaks that needs to be investigated. Mean loan amount is 8337$


```{r echo=FALSE, fig.width=10, fig.height=6}
summary(df_s$LoanStatus)

ggplot(data=df_s, aes(df_s$LoanStatus)) + 
  geom_histogram(col="black", alpha = .6, stat="count") + 
  labs(title="Histogram for Loan Status") +
  labs(x="Loan Status", y="Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


Largest count of loans is still taged as current. From the past due loans 
the largest cathegory is past due between 1-15 days.


```{r echo=FALSE, fig.width=10, fig.height=6}

summary(df_s$DebtToIncomeRatio)

ggplot(data=df_s, aes(df_s$DebtToIncomeRatio)) + 
  geom_histogram(binwidth = 0.05,
                 col="black", 
                 alpha = .6,
                 aes(fill=..count..)) + 
  labs(title="Histogram for DebtToIncomeRatio") +
  labs(x="DebtToIncomeRatio - log scale", y="Count") +
  scale_fill_gradient("Count", low = "green", high = "red") +
  xlim(0, 2) +
  scale_x_log10() 

### calculate ratio of customers that have more than 4 times income 
## to debt ratio
df_s %>% 
  select(DebtToIncomeRatio) %>% 
  filter(DebtToIncomeRatio > 4) %>% 
  summarise(count= n()) %>% 
  mutate(ratio_4_times = (count / 105383) * 100)

```

Mean debth to income ratio is 0.28 and maximum is 10.010.  The count to loans
that have ratio more than 4 times compared to all loans (minus NA s) is 0.35%.

```{r echo=FALSE}

#plot
ggplot(df_s, aes(factor(Term/12))) +
  geom_bar() +
  labs(title="Histogram for Term Length in Years") +
  labs(x="Term Length [Years]", y="Count") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

```

Length of loans are distributed in 3 cathegories, 
1 3 and 5 year loans where count is 3>5>1 years.



```{r echo=FALSE, fig.width=10, fig.height=6}
summary(df_s$BorrowerRate)

#plot
ggplot(data=df_s, aes(df_s$BorrowerRate)) + 
  geom_histogram(col="black", 
                 alpha = .6,
                 aes(fill=..count..)) + 
  labs(title="Histogram for Borrowers Interest Rates per entire loan") +
  labs(x="Borrower Interest Rate [%]", y="Count") +
  scale_fill_gradient("Count", low = "green", high = "red") 

```

Distribution of Borrowers interest loans is quite narrow with a mean at 0.19%.
Distribution is a slightly skewed to the left.
The values it self seem a bit suspiceous. I would expect to see interest rates
at least one magnitude higher from 0 to 5 percent so it could be an issue with
the original dataset. I real case scenario I would have a direct dialogue with
a domain expert in the topic:)


```{r echo=FALSE, fig.width=10, fig.height=6}
summary(df_s$IncomeRange)

df_s$IncomeRange <- factor(df_s$IncomeRange, 
                           c("$0", "$1-24,999",
"$25,000-49,999", "$50,000-74,999", 
"$75,000-99,999", "$100,000+","Not displayed", "Not employed"))

ggplot(data=df_s, aes(df_s$IncomeRange)) + 
  geom_histogram(col="black", alpha = .6, stat="count") + 
  labs(title="Bar chart for Income Range") +
  labs(x="Income Range [USD]", y="Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

The highest income range for the borrowers is in two cathegories which in total
range between 25 000 and 74 999 us dollars.

```{r echo=FALSE, fig.width=10, fig.height=6}
summary(df_s$EmploymentStatus)
names(df_s)
ggplot(data=df_s, aes(reorder(EmploymentStatus,EmploymentStatus,
                     function(x)-length(x))))  + 
  geom_histogram(col="black", alpha = .6, stat="count") + 
  labs(title="Bar chart for Employement status") +
  labs(x="Employement Status", y="Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

90% of the borrowers that take loans are either Employed, Full-time eployed or 
Self-Employed

```{r echo=FALSE, fig.width=10, fig.height=6}

ggplot(data=df_s, aes(reorder(Occupation,Occupation,
                     function(x)-length(x)))) + 
  geom_histogram(col="black", alpha = .6, stat="count") + 
  labs(x="Occupation", y="Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
```

Top 5 occupations of borrowers were as following: Professional, Computer 
programmer, Executive, Teacher and Administrative assistant.
Bottom of the list is dominated by students.


```{r echo=FALSE, fig.width=10, fig.height=6}
summary(df_s$IsBorrowerHomeowner)

##plot
ggplot(data=df_s, aes(reorder(IsBorrowerHomeowner,IsBorrowerHomeowner,
                     function(x)-length(x)))) + 
  geom_bar(col="black", alpha = .6, stat="count") + 
  labs(x="IsBorrowerHomeowner", y="Count") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

  
```  

In the entire dataset number of homeowners vs the ones that do not own a home
is very similar (False- 56459 vs True - 57478).


```{r echo=FALSE, fig.width=10, fig.height=6}
##extracting years from LoanOriginationDate
df_s <- df_s %>% 
  mutate(LoanOriginationYear = (substring(df_s$LoanOriginationDate,0,4)))
##plot
ggplot(data=df_s, aes(LoanOriginationYear))+ 
  geom_bar(col="black", alpha = .6, stat="count") + 
  labs(x="LoanOriginationYear", y="Count") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))
```  

Distribution of years when the loan were originated. Note the peak in 2013.


```{r echo=FALSE, fig.width=10, fig.height=6}
# create lookup table
listingcode = c(0, 1, 2,  3,  4,  5,  6,  7,  8,  9, 10, 11, 12, 13, 14, 15, 
                16, 17, 18, 19, 20) 
listingname = c("Not Avilable", "Debt Consolidation", "Home Improvement", 
                "Business", "Personal Loan", 
           "Student Use", "Auto", "Other", "Baby and Adoption", "Boat",
           "Cosmetic Procedure", "Engagement Ring", "Green Loans",
           "Household Expenses", "Large Purchases","Medical and Dental",
           "Motorcycle","RV", "Taxes", "Vacation", "Wedding Loans")
lookup = data.frame(listingcode = as.integer(listingcode), 
                    ListingCathegory = listingname)


## Map a lookup table to the df_s
df_s <- df_s %>%
   left_join(lookup, by= c("ListingCategory..numeric." ="listingcode")) 


## plot
ggplot(data=df_s, aes(reorder(ListingCathegory,ListingCathegory,
                     function(x)-length(x)))) + 
  geom_histogram(col="black", alpha = .6, stat="count") + 
  labs(title="Bar chart for ListingCathegory") +
  labs(x="ListingCathegory", y="Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
  

```

Most common listing cathegory for a loan are Debt consolidation,
Home improvement and Business, if we exclude Not available and other.


# Univariate Analysis

### What is the structure of your dataset?

Subset of the data consists of 15 variables and is spread in 113937 rows.
Data completeness report gave us an insight that most of the observations have
values apart from the ClosedDate which is missing in 50% of the cases, and
that is justifiable since it applies only to Cancelled, Completed, Chargedoff 
and Defaulted loan statuses.

I didn t manage to understand the root cause for the 871 of duplicated ListingKeys.
 

### What is/are the main feature(s) of interest in your dataset?
Main feature of interest in the dataset is the amount of loan.

### What other features in the dataset do you think will help support your 
### \ Did you create any new variables from existing variables in the dataset?
Supporting features that I brought in externally are related to mapping out
the US state codes into a human language. I did the same with the 
listing cathegories but  remapping was done from an additional
explanatory excel spreadsheet.
Spatial context - geometry of the states would be of hel to see spatial 
distribution of loans and loan related variables.


### Of the features you investigated, were there any unusual distributions? 
### \ Did you perform any operations on the data to tidy, adjust, or change the 
### form \ of the data? If so, why did you do this?
Form of the data was not changed, i.e. data is in tidy format.


# Bivariate Plots Section

```{r echo=FALSE, fig.width=10, fig.height=6, essage=FALSE, warning=FALSE}

ggplot(df_s, aes(factor(Term/12),LoanOriginalAmount, fill=Term)) +
  labs(title="Box Plot") +
  labs(x="Term length for loans [years]", y="Loan Amount [USD]") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(legend.position="none") +
  geom_jitter( alpha = .01)  + 
  geom_boxplot( alpha = .5,color = 'blue')+ 
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 3)


```

There is a positive relationship when comparing median as a metric between
Loan amounts and 1, 3 and 5 years as lenght of loan. Higher to Term,
higher the median loan amount.

```{r echo=FALSE, fig.width=10, fig.height=6, message=FALSE, warning=FALSE}
## Summarizing original amounts of loans per listing cathegory
MeanPerCathegory <- df_s %>% 
  group_by(ListingCathegory) %>% 
  summarise(mean = mean(LoanOriginalAmount)) %>% 
  arrange(desc(mean))

#plot
fill <- "#4271AE"
line <- "#1F3552"
ggplot(df_s, aes(x=ListingCathegory, y=LoanOriginalAmount)) +
  geom_boxplot(fill = fill, colour = line, alpha = 0.7,
                     outlier.colour = "#1F3552", outlier.shape = 20) +
  labs(x="ListingCathegory", y="LoanOriginalAmount[USD]")+
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  theme(legend.position="none")  +
  geom_jitter( alpha = .01)  + 
  geom_boxplot( alpha = .5,color = 'blue')+ 
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 3) 
 
```

Top 3 highest loans on average are targeting Debt consolidation, baby and 
adoption,and business, on the other side of the spectrum top 3 lowest loans 
on average are targeting student use, personal loans and buying an automobile.

```{r echo=FALSE, fig.width=10, fig.height=6,  message=FALSE, warning=FALSE}
## summarizing means of original loan amounts per occupation groups
## filtered datasets for means over 10 000USD
MeanLoanPerOccupation <- df_s %>% 
  group_by(Occupation) %>% 
  summarise(mean = mean(LoanOriginalAmount)) %>% 
  filter(mean>10000)

# Plot
theme_set(theme_bw())
ggplot(MeanLoanPerOccupation, aes(reorder(Occupation,mean),mean)) + 
  geom_point(size=3) + 
  geom_segment(aes(x=Occupation, 
                   xend=Occupation, 
                   y=0, 
                   yend=mean)) + 
  labs(title="Lollipop Chart", 
       subtitle="Occupation Vs Mean Loan") + 
  labs(x="Occupation", y="Mean Loan Amount [USD]") +
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) +
    coord_flip()

```

Top 5 on average highest loans are initiated by folowing occupation groups:
Judges, Pharmacists and Doctors (dataset filtered on more than 10 000USD loans).

```{r echo=FALSE, fig.width=10, fig.height=6, message=FALSE, warning=FALSE}

#MeanInterestsPerLoanStatus <- df_s %>% 
 # group_by(ListingCathegory) %>% 
  #summarise(mean = mean(LoanOriginalAmount)) %>% 
  #arrange(desc(mean))

## plot
ggplot(df_s, aes(x=LoanOriginalAmount, y=DebtToIncomeRatio)) +
  geom_jitter(alpha=0.05,aes()) +
  geom_smooth() +
  scale_y_log10() +
  labs(x="Original Loan Amount [USD]", y="Debt to Income Ratio") 

  #theme(legend.position="none")  +
  # scale_colour_gradientn(colours=rainbow(4)) +
 # geom_encircle(aes(x=LoanOriginalAmount, y=DebtToIncomeRatio), 
  #              data=df_s_retired, 
   #             color="red", 
    #            size=2, 
     #           expand=0.08)

#df_s_retired <- df_s %>% 
 # filter(EmploymentStatus=="Retired")

```

What I expected to see when plotting Original Loan amount against depth to 
to income ratio is more negative correlation. What is clear in this plot iz that
if we exclude >10 debt to income ratio, is that in a range of 1 to 10 of 
debt to income ratio there is very few data point in the space between
10 000 and 30 000 when comparing it to below 10 000 where the majority of the 
points lay i.e. burrowers with higher DTI ratio take lower loans.


```{r echo=FALSE, fig.width=10, fig.height=6, message=FALSE, warning=FALSE}

## combining distinct combinations of Occupation groups and income range
## groups, counting them and keeping only counts above 2000.
MeanIncomePerOccupation <- df_s %>% 
  group_by(Occupation,IncomeRange) %>% 
  tally() %>% 
  mutate(OccupationPlusIncome = paste(Occupation,"AND", IncomeRange)) %>% 
  filter(n>2000)

#plot
theme_set(theme_bw())
ggplot(MeanIncomePerOccupation, aes(reorder(OccupationPlusIncome,n),n)) + 
  geom_point(size=3) + 
  geom_segment(aes(x=OccupationPlusIncome, 
                   xend=OccupationPlusIncome, 
                   y=0, 
                   yend=n)) + 
  labs(title="Lollipop Chart", 
       subtitle="OccupationPlusIncome Vs Count") + 
  labs(x="OccupationPlusIncome", y="Count") +
  theme(axis.text.x = element_text(angle=80, vjust=0.6)) +
    coord_flip()

```

Highest count of initiated loans in the datacomes from Professionals 
in the income groups: 50,000-74,999USD and 100,000+ USD (if we exclude 
combinations defined as Other).

```{r echo=FALSE, fig.width=10, fig.height=6, message=FALSE, warning=FALSE}


#plot
fill <- "#4271AE"
line <- "#1F3552"
ggplot(df_s, aes(x=LoanOriginationYear, y=BorrowerRate)) +
  geom_boxplot(fill = fill, colour = line, alpha = 0.7,
                     outlier.colour = "#1F3552", outlier.shape = 20) +
  labs(x="Loan Originated [Year]", y="Borrower Interest Rate for the Loan [%]")+
  geom_jitter( alpha = .01)  + 
  geom_boxplot( alpha = .5,color = 'blue')+ 
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 3) 

  


```


It is evident that median BorrowerRate has been increasing over the years with
a peak in 2011 and dropping again towards 2014. In addition there is a trend
in narrowing IQR from 2010 onwards.

```{r echo=FALSE, fig.width=10, fig.height=6, message=FALSE, warning=FALSE}

#plot
fill <- "#4271AE"
line <- "#1F3552"
ggplot(df_s, aes(x=LoanOriginationYear, y=LoanOriginalAmount)) +
  geom_boxplot(fill = fill, colour = line, alpha = 0.7,
                     outlier.colour = "#1F3552", outlier.shape = 20)  +
  labs(x="Loan Originated [Year]", y="Original Loan Amount [USD]")  +
  geom_jitter( alpha = .01)  + 
  geom_boxplot( alpha = .5,color = 'blue')+ 
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 3)

```

Medians for loan original amounts have been increasing since 2009 with a peak
in 2014. In addition, there has almost not been outliers in a range over 
30 000USD until year 2013 and 2014.



# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the 
### investigation. How did the feature(s) of interest vary with other 
### features in the dataset? 

Main obesrvations:

  * positive relationship between increase of loan amounts medians and
  increase of length of loans - not a surprise
  
  * Medians for loan original amounts have been increasing since 2009 with a peak
  in 2014. In addition, there has almost no outliers in a range over 
  30 000USD until year 2013 and 2014.
  
  * It is evident that median BorrowerRate has been increasing over the years 
  with a peak in 2011 and dropping again towards 2014. In addition there is 
  a trend in narrowing IQR from 2010 onwards.

  * Top 5 on average highest loans are initiated by folowing occupation groups:
  Judges, Pharmacists and Doctors (dataset filtered on more than 
  10 000USD loans).
  
  * Top 3 highest loans on average are targeting Debt consolidation, baby and 
  adoption and business, on the other side of the spectrum top 3 lowest 
  loans on average are targeting student use, personal loans and buying 
  an automobile.



### Did you observe any interesting relationships between the other features 
### \(not the main feature(s) of interest)?

Highest count of initiated loans in the datacomes from Professionals 
in the income groups: 50,000-74,999USD and 100,000+ USD (if we exclude 
combinations defined as Other)


### What was the strongest relationship you found?
THere is no in particular strong relationship found appart from the obvious ones
which relate to OriginalLoanAmounts and levels of income, higher paid professions,
increase in term length.
I would benefit of having more in depth domain knowledge in order to scrutinize
some of the more intricate relationships.

# Multivariate Plots Section


```{r echo=FALSE, fig.width=16, fig.height=10, Multivariate_Plots}


MeanLoanPerOccupationPerYear <- df_s %>% 
  group_by(Occupation,LoanOriginationYear,IsBorrowerHomeowner) %>% 
  summarise(meanLoanOriginalAmount = mean(LoanOriginalAmount)) 

ggplot(MeanLoanPerOccupationPerYear, aes(x=LoanOriginationYear, 
                                         y=meanLoanOriginalAmount)) + 
  geom_point(aes(col=IsBorrowerHomeowner),size=3) + 
  facet_wrap(~Occupation) +
   theme(axis.text.x = element_text(angle=80, vjust=0.6)) +
   geom_hline(yintercept = 5000, color="black",size=0.5)  +
  labs(x="Loan Originated [Year]", y="Mean Original Loan Amount [USD]")

```

There is a lot of information in this graph which summarizes mean loans grouped
per year faceted per Occupation and colored by a flag if borrower is a homeowner 
or not. 
What can be seen at a first glance is that means are either equal or bigger if
a borrower is a homeowner.
Moreover, means are increasing in general regardless of occupation over 
the period of last 5 years.
Horisonal line on the graph is highlighting mean of 5000USD loans and by 
focusing on points below that line we can see that they belong to a Student 
groups which doesn't seem to change over time significantly and colors are 
associated to a flag - is not a homeowner.



```{r echo=FALSE, fig.width=16, fig.height=10}

MeanLoanPerListingCathegoryPerYear <- df_s %>% 
  group_by(ListingCathegory,LoanOriginationYear,IsBorrowerHomeowner) %>% 
  summarise(meanLoanOriginalAmount = mean(LoanOriginalAmount)) 

## plot
ggplot(MeanLoanPerListingCathegoryPerYear, aes(x=LoanOriginationYear, 
                                               y=meanLoanOriginalAmount)) + 
  geom_point(aes(col=IsBorrowerHomeowner), size=3) + 
  facet_wrap(~ListingCathegory) +
    theme(axis.text.x = element_text(angle=80, vjust=0.6))  +
  labs(x="Loan Originated [Year]", y="Original Loan Amount [USD]") 

```

Along similar lines, if a borrower is a homeowner, he or she will have on average
similar or higher loans than if a borrower is not a homeowner.
It is interesting to see that some of the ListingCathegories are constrained to
last 4 years and some are more expanding over almost entire time span of a 
dataset, like debt consolidation, home improvements, businesses and automobiles.
In addition, personal loans and student use is absent in last 4-6 years.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the 
### \investigation. Were there features that strengthened each other 
### in terms of \looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

There is a lot of information in two graphs in multivariate part which 
summarizes mean loans grouped per year faceted per Occupation and colored 
by a flag if borrower is a homeowner or not. 
What can be seen at a first glance is that means are either equal or bigger if
a borrower is a homeowner.
Moreover, means are increasing in general regardless of occupation over 
the period of last 5 years.
Horisonal line on the graph is highlighting mean of 5000USD loans and 
by focusing on  points below that line we can see that they belong to
a Student groups which doesn't seem to change over time significantly 
and colors are associated to a flag - is not a homeowner.

Along similar lines, if a borrower is a homeowner, he or she will have on average
similar or higher loans than if a borrower is not a homeowner.
It is interesting to see that some of the ListingCathegories are constrained to
last 4 years and some are more expanding over almost entire time span of a 
dataset, like debt consolidation, home improvements, businesses and automobiles.
In addition, personal loans and student use is absent in last 4-6 years.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, fig.width=10, fig.height=6, message=FALSE, warning=FALSE, Plot_One}

ggplot(data=df_s, aes(df_s$LoanOriginalAmount)) + 
  geom_histogram(breaks=seq(0, max(df_s$LoanOriginalAmount), by = 1000), 
                 col="black", 
                 alpha = .6,
                 aes(fill=..count..)) + 
  labs(title="Histogram of the Original Loan Amounts in USD") +
  labs(x="Original Loan Amount [USD]", y="Count") +
  scale_fill_gradient("Count", low = "green", high = "red") 


```

### Description One
Main feature of my interest was the maginute of original loan amounts
Histogram shows count of the magnitude or the original loan amounts
that borrowers have taken. What is evident from this plot is the skewness of
the distribution with the right tail. 
In addition, There are several local peaks breaking the ideally
right skewed distribution.

### Plot Two
```{r echo=FALSE, fig.width=10, fig.height=6, message=FALSE, warning=FALSE, Plot_Two}
fill <- "#4271AE"
line <- "#1F3552"
ggplot(df_s, aes(x=LoanOriginationYear, y=LoanOriginalAmount)) +
  geom_boxplot(fill = fill, colour = line, alpha = 0.7,
                     outlier.colour = "#1F3552", outlier.shape = 20) +
  labs(title="Box Plot of Original Loan Amounts  vs Year Loan was initiated") +
  labs(x="Initiation Year of the Loan", y=" Original Loan Amount [USD]")  +
  geom_jitter( alpha = .01)  + 
  geom_boxplot( alpha = .5,color = 'blue')+ 
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 3)
```

### Description Two
Plot two is focusing on showing a relationship of distributons of initial loan
amounts vs year the loan is initiated.
Medians for loan original amounts have been increasing since 2009 with a peak
in 2014. In addition, there has almost absence of the outliers in a range over 
30 000USD until year 2013 and 2014.
All of the distributions are skewed.


### Plot Three
```{r echo=FALSE, fig.width=10, fig.height=6, message=FALSE, warning=FALSE, Plot_Three}

## summarizing means for groups ListingCathegory,LoanOriginationYear,
## IsBorrowerHomeowner
MeanLoanPerListingCathegoryPerYear <- df_s %>% 
  group_by(ListingCathegory,LoanOriginationYear,IsBorrowerHomeowner) %>% 
  summarise(meanLoanOriginalAmount = mean(LoanOriginalAmount)) 

## plot
ggplot(MeanLoanPerListingCathegoryPerYear, aes(x=LoanOriginationYear, 
                                               y=meanLoanOriginalAmount)) + 
  geom_point(aes(col=IsBorrowerHomeowner), size=3) + 
  facet_wrap(~ListingCathegory) +
    theme(axis.text.x = element_text(angle=80, vjust=0.6))  +
  labs(title="Scatter Plot of Average Original Loan Amounts vs Year Loan was 
       Initiated") +
  labs(x="Initiation Year of the Loan", y="Average Original Loan Amount [USD]") 
```

### Description Three
Plot three carries a lot of information which enriches the plots from first and
second plot with additional variables to see multivariate relationships with
the main feature of interest - average original loan amounts per year, 
faceted based on the listing cathegory and colored by a variable which
says if the borrower is already a homeowner or not.

What can be deduced is the following, if a borrower is a homeowner,
he or she will have on average
similar or higher loans than if a borrower is not a homeowner.
It is interesting to see that some of the ListingCathegories are constrained to
last 4 years and some are more expanding over almost entire time span of a 
dataset, like debt consolidation, home improvements, businesses and automobiles.
In addition, personal loans and student use are completely absent in last 4-6 
years.

------

# Reflection

Exploration of dataset was challenging because firstly due to the time
constraints I had to pick and choose 10-15 variables from 81 of them which
was not easy not having extensive domain knowledge on the topic.

Univariate analyses gave me room to explore distributions 
of most important variables in the dataset.

Bivariate analyses pointed me in directions where should I focus when exploring
two variables at the same time and increased my curiousity to explore timeseries
change.

Multivariate analyses in fully made me able to explore the relationships between
time, magnitude of loans and two additional variables like home ownership
and listing type in the same visualisation - very powerful tools.

Some of the strugles I encountered were lack of domain knowledge in the topic
and wrangling the datasets, preparing the data so it is in right format
for plotting to happen in the right manner.

What went pretty well is generally how I handeled the who EDA process:)

As a future insight I would like to explore the dataset more in spatial domain
and also start including more relevant variables from the full dataset.

In addition I would like to spend more time into quantifying potential 
relationships that were detected during the visual inspections of plots.



### A list of Web sites, books, forums, blog posts, github repositories:

http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html

https://www.growingfamilybenefits.com/credit-scores-interest-rates-relationship/

https://stackoverflow.com/questions/15625990/how-to-set-size-for-local-image-using-knitr-for-markdown

http://t-redactyl.io/blog/2016/04/creating-plots-in-r-using-ggplot2-part-10-boxplots.html

https://stackoverflow.com/questions/30023610/how-to-plot-2-categorical-variables-on-x-axis-and-two-continuous-variables-as-f

https://admccarthy.github.io/Red_Wines/

https://docs.google.com/document/d/1-f3wM3mJSkoWxDmPjsyRnWvNgM57YUPloucOIl07l4c/pub